PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-06-23 manuscript_00.Rmd
new plots

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
data_dir_vGiga <- '~/data/scEiaD_gigascience/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
c_d <- plot_grid(c,d, rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))

@@

plot_grid(a_aa_b, c_d, proportion_bar_plot, ncol = 3, rel_widths = c(1,1.6, 3))
```

@@

```{r}
# fix  occasional NA
# data made from src/ensdb.R
diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
diff_tab <- diff_tab_big %>% 
  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))

@@

```{r, fig.height=9, fig.width=7}
g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=5, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)

@@

## Tease apart closely related cell types
```{r}
x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Base %in% c("Early RPC", "Late RPC", "RPC")) %>%  
  filter(Group == 'CellType (Predict)', 
         Against %in% c("Early RPC", "Late RPC", "RPC")) %>% 
  group_by(Base, Against, Gene) %>% 
  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-3)) %>% 
  arrange(-`mean log2FC`) %>%   
  collect() %>% 
  filter(!is.na(`mean log2FC`))

  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
  left_join(gene_tab, by = 'GENEID') %>% 
  relocate(SYMBOL, GENEID) %>% 
  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
  arrange(-`mean log2FC`)
```

@@

```{r}
# flip flop diff genes
down_tab <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) < -2) %>% 
  collect() %>% 
  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% 
  mutate(k = paste(Symbol, Base, sep = '_'))

diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
  filter(k %in% down_tab$k, Base %in% c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )) %>% 
  arrange(padj)

diff_tab_big %>% 
  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
  group_by(Symbol, Base, Organism) %>% 
  summarise(Count = n()) %>% 
  filter(Count > 8, Base == 'Rod') %>% arrange(-Count)
```

```{r diff_difff_exp_genes, fig.height = 3, fig.width = 8, echo = F, message = F}
source('src/make_exp_plot.R')

input <- list()
input$exp_plot_groups <- c('organism')
input$exp_plot_facet <- c('CellType_predict')
input$exp_plot_genes <- c('SCRG1 (ENSG00000164106)')
input$exp_filter_cat <- 'CellType_predict'
input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
input$exp_plot_col_num <- 16
input$exp_filter_min_cell_number <- 50

make_exp_plot(input, scEiaD_2020_v01, meta_filter)
```

@@

input <- list()
input$exp_plot_groups <- c('organism')
input$exp_plot_facet <- c('CellType_predict')
input$exp_plot_genes <- c('RHO (ENSG00000163914)')
input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
input$exp_plot_col_num <- 16
input$exp_filter_min_cell_number <- 50

>>>>

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
c_d <- plot_grid(c,d + theme(legend.position = "none"), rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))

@@

e <- plot_grid(proportion_bar_plot, labels = c('d'))

plot_grid(a_aa_b, c_d , e, ncol = 3, rel_widths = c(1,1.6, 2.2))
```

@@

```{r}
# fix  occasional NA
# data made from src/ensdb.R
diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
diff_tab <- diff_tab_big %>% 
  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))

@@

```{r, fig.height=11, fig.width=6}
g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)

@@

## Tease apart closely related cell types

Neurogenic and RPC have very similar expression patterns. We leverage our more specific differential test to directly assess genes were are different between neurogenic cells and RPC. We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.
```{r}
x <-  scEiaD_2020_v01 %>% 
    tbl("diff_testing") %>% 
    filter(Against  == "RPC",   
           Group == 'CellType (Predict)', 
           Base == "Neurogenic Cell") %>% 
  group_by(Base, Against, Gene) %>% 
  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
  arrange(-`mean log2FC`) %>%   
  collect() %>% 
  filter(!is.na(`mean log2FC`))

  # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
  # left_join(gene_tab, by = 'GENEID') %>% 
  # relocate(SYMBOL, GENEID) %>% 
  # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
  # arrange(-`mean log2FC`)

down <- x %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0) %>% head(5) %>% pull(Gene)
up <-  x %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(5) %>% pull(Gene)
source('src/make_exp_plot.R')

input <- list()
input$exp_plot_groups <- c('study_accession')
input$exp_plot_facet <- c('CellType_predict')
input$exp_plot_genes <- down
input$exp_filter_cat <- 'CellType_predict'
#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
input$exp_filter_on <- c("Neurogenic Cell", "RPC" )
input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
#input$exp_plot_ylab <- '% of Cells Detected'
input$exp_plot_col_num <- 16
input$exp_filter_min_cell_number <- 50
input$flip_facet <- TRUE

a <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)

input$exp_plot_genes <- up
b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) 
plot_grid(a,b, labels = c('a','b'))
```

@@

```{r, fig.width=3, fig.height=7}
# flip flop diff genes
down_tab <- down_tab_big %>% 
   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
  filter(Base %in% ws_oct, 
         Against == 'All')

org_diff_diff_genes <- diff_tab_big %>% 
  mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
  filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
         Against == 'All') %>% 
  arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')'))

genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
test <- scEiaD_2020_v01 %>%
  tbl("diff_testing") %>%
  filter(Gene %in% genes) %>%
  collect() %>%
  mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
  mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
  filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
         Group == 'CellType (Predict)',
         Against == 'All') %>%
  select(Gene, Organism, log2FoldChange, Base)

x <- test %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
organism <- colnames(x)[c(3,4,5)]
x <- x %>% data.frame()
row.names(x) <- paste0(x$Gene,'_', x$Base)

x <- x[,-c(1,2)]

genes <- row.names(x) %>% gsub(' \\(.*','',.)

base_yo <- row.names(x) %>% gsub('.*_','',.)

hm <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
                              row_split = base_yo, show_row_dend = FALSE, 
                              row_labels = genes, 
                              column_title_rot = 90, 
                              name = 'log2FC', use_raster = TRUE,
                              row_title_rot = 0)
hm
```

```{r}
library(enrichR)
dbs <- c('GO_Biological_Process_2018', 'GO_Molecular_Function_2018')
enriched_cone <- enrichr(org_diff_diff_genes %>% filter(Base == 'Rod') %>% pull(Symbol), dbs)

@@

input <- list()
input$exp_plot_groups <- c('organism')
input$exp_plot_facet <- c('CellType_predict')
input$exp_plot_genes <- c('RHO (ENSG00000163914)')
input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
input$exp_plot_col_num <- 16
input$exp_filter_min_cell_number <- 50
input$flip_facet <- FALSE
